---
title: "Data Science for Public Policy Final Project"
subtitle: "Examining the Relationship Between Economic Status and Public Health Outcomes in the D.C. Area"
author: "Adriana Vance, Kyle Kim, Elena Koshkin, Kamiryn Rose"
execute:
  warning: false
format:
  html:
    embed-resources: true
---

## Data Wrangling and Exploratory Data Analysis

Loading necessary packages
```{r message=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)
library(tidycensus)
library(purrr)
library(sf)
library(dplyr)
library(patchwork)
library(recipes)
library(ggplot2)
library(tidyclust)
library(tigris)
library(broom)
library(factoextra)
library(tmap)
library(mapproj)
library(leaflet)
library(randomForest)
library(caret)
library(tidymodels)
library(hexbin)
```

<<<<<<< HEAD
## Part 01
### Data Wrangling and Plots
```{r echo = false, message=FALSE, warning=FALSE}
#Read in CSV of CDC data
Health <- read_csv("PLACES.csv")

#### Pulling geospatial data for mapping purposes 
#Loading Tigris Shapefiles from the census year that matches the geometry of our health and economic data (2010). These shapefiles will come in handy when mapping health outcomes and economic data later in the project.

#pulling tigris package tracts data

Maryland_tracts <- tracts(
state = "MD",
county = c("Montgomery", "Prince George's"),
cb = FALSE,
resolution = "500k",
year = 2010)

Virginia_tracts <- tracts(
  state = "VA",
  county = "Fairfax County",
  resolution = "500k", 
  year = 2010
)

DC_tracts <- tracts(
  state = "DC",
  resolution = "500k",
  year = 2010
)

DMV_tracts<- rbind_tigris(DC_tracts, Virginia_tracts, Maryland_tracts)

```

### Health Outcomes Data
The CDC PLACES dataset uses small-area estimation methods to provide model-based mapping of 36 different chronic disease-related measures throughout the United States. These measures include 13 health outcomes, 9 prevention practices, 4 health risk behaviors, 7 disabilities, 3 health statuses, used to estimate the rates of various chronic conditions by census tract level. The CDC PLACES data was downloaded from the CDC website and sifted to select data from exclusively DC, Maryland, and Virginia. Please download the CSV file provided, "PLACES.csv", to load the correct dataframe. 

```{r echo = false, message=FALSE, warning=FALSE}
#Read in CSV of CDC data
Health <- read_csv("PLACES.csv")

# Look at types of health outcome measures
Measures <- Health %>%
  group_by(Measure) %>%
  summarize(n = n())

MeasureIDs <- Health %>%
  group_by(MeasureId) %>%
  summarize(n = n())

# We want to look for current rates of asthma, heart disease, diabetes, stroke, and depression
Health <- Health %>%
  filter(
    MeasureId == "CASTHMA" |
      MeasureId == "CHD" |
      MeasureId == "DIABETES" |
      MeasureId == "STROKE" |
      MeasureId == "DEPRESSION")

```

We chose asthma, coronary heart disease, stroke, and depression as our health outcomes of interest because research shows a strong connection between these conditions and economic status. The most useful observations from this dataset will be MeasureId, which is the type of health outcome, and Data_Value, which is the percentage of the total population of a given census tract that report said health outcome. 

#### Health Outcomes Distribution Map

```{r}
# Change geometry to be readable by SF

Health_map <- Health %>%
  st_as_sf(wkt = "Geolocation", remove = FALSE) %>%
  st_set_crs(value = 4326) # setting CRS


# Map health outcomes
Health_map %>%
  ggplot() +
  geom_sf(data = DMV_tracts) +
  geom_sf(aes(color = MeasureId)) +
  labs(
    title = "Health Outcomes in the DMV",
    subtitle = "Distribution by Census Tract",
    caption = "CASTHMA = Current Asthma\nCHD = Coronary Heart Disease",
    color = "Health Measures ID"
  )

```
This map demonstrates that at least one of the health outcomes we selected are present in each census tract in our dataset. Later in the project, we will dive further into the connection between location and health outcomes, and use economic data to further understand how health outcomes differ in each census tract. There isn't an obvious pattern among the tracts that show a concentration of specific health outcomes by specific area.


### Debt-to-Income Ratio Data
The debt-to-income ratio dataset comes from the Federal Housing Finance Agency which records data about home purchase loans. This dataset has information on individual loans taken out for single-family homes, along with specification of census tracts. The data also include the debt-to-income ratio of each borrower, which is an indication of how much debt the borrower has taken on in relation to their income. The debt-to-income ratio (or DTI) is generally regarded as a reliable indicator of a borrower’s financial health, or ability to pay monthly debt payments. Other useful variables in this dataset that will be used in our analysis are census tract indicators, the percentage of minorities in a given census tract, tract median income, and mortgage amount. 

We chose to download the dataset from 2021, as this will reflect the same year as our CDC health outcomes dataset. To download the correct dataset for our analysis, go to this [link:](https://www.bcbs.com/the-health-of-america/articles/healthy-communities-mean-better-economy) and download the ZIP file for the 2021 Freddie Mac dataset. Alternatively, you can use the txt file we included with our submission titled "debt_to_income.txt"

```{r echo = FALSE, warning = FALSE, message = FALSE}
# loading debt to income ratio data

debt_to_income <- read.table("debt_to_income.txt", header = FALSE) 
# Have to manually name columns

colnames(debt_to_income) <- c('enterprise', 'record_num','state_postal_code', 'MSA', 'COUNTYFP', 'NAME10', 'percent_minority', 'median_income', 'MSA_median_income', 'tract_income_ratio', 'borrower_income','AMI', 'borrower_income_ratio', 'UPB','purpose_of_loan', 'fed_guarantee', 'num_borrowers', 'first_time_borrower', 'borrower_race_1', 'race_2', 'race_3', 'race_4', 'race_5', 'borrower_ethnicity', 'co_borrower_race_1', 'co_race_2', 'co_race_3', 'co_race_4', 'co_race_5',   'co_borrower_ethnicity', 'borrower_gender', 'co_borrower_gender', 'borrower_age', 'co_borrower_age', 'occupancy_code', 'rate_spread', 'HOEPA_status', 'property_type', 'lien_status', 'borrower_over_sixtytwo', 'co_borrower_over_sixtytwo', 'LTV_ratio', 'note_date', 'mortgage_term', 'num_units', 'interest_rate', 'note_amount', 'preapproval', 'application_channel', 'AUS_name', 'borrower_credscore_model', 'co_borrower_credscore_model', 'DTI_ratio', 'discount_points', 'rate_period', 'manufactured_home_ownership', 'property_value', 'rural_tract', 'lower_MS_delta', 'Mid_Appalachia', 'persistent_poverty_county', 'concentrated_poverty', 'high_opportunity_area', 'QOZ')

counties <- c(1, 59, 31, 33) # Name county vector to filter dataset to needed counties only 

debt_to_income <- debt_to_income %>%
  filter(MSA == 47900) 

debt_to_income <- debt_to_income %>%
  filter(COUNTYFP == counties) # Filtering to needed counties 

# Mutate dataset to match county names of TIGRIS shapefile
debt_to_income <- debt_to_income %>%
  mutate(COUNTYFP = 
    if_else(COUNTYFP == 1, "001", as.character(COUNTYFP)))

debt_to_income <- debt_to_income %>%
  mutate(COUNTYFP = 
           if_else(COUNTYFP == 59, "059", as.character(COUNTYFP)))
    
debt_to_income <- debt_to_income %>%
  mutate(COUNTYFP = 
           if_else(COUNTYFP == 31, "031", as.character(COUNTYFP)))

debt_to_income <- debt_to_income %>%
  mutate(COUNTYFP = 
           if_else(COUNTYFP == 33, "033", as.character(COUNTYFP)))
```
The debt-to-income ratio dataset did not come with column names, so we used the data documentation to manually input the names. Some of the column names were changed to match the tigris shapefiles which will be used in later mergers.

### Exploratory Data Analysis using the debt-to-income ratio dataset
```{r echo=FALSE, warning=FALSE, message=FALSE}

# Exploratory data analysis to better understand the data 
debt_totals <- debt_to_income %>%
  summarise(n = n())
  
debt_to_income %>%
  filter(purpose_of_loan == 3 | 4 | 5 | 6) %>%
  ggplot() + 
  geom_bar(mapping = aes(x = purpose_of_loan)) + 
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7), minor_breaks = NULL, labels = c("Purchase", "Refinancing", " ", "Home Improvement", " ", " ", "Refinancing (cash-out)")) +
  ggtitle("Count of borrowers by loan type") 
  
```
Here we can see that most of the loans taken out are for the purposes of refinancing a mortgage, while a smaller number are for initial purchase of homes. 

```{r}
debt_to_income %>%
  ggplot() + 
  geom_hex(mapping = aes(x = percent_minority, y = note_amount)) +
    scale_fill_continuous(low = "darkgreen", high = "yellow") + 
  scale_y_continuous() 


```
This hexbin plot demonstrates that there is a concentration of loans among the 300,000 dollar amount in 30-40% minority census tracts. There is also a significant concentration of single family home loans in the near-100% minority census tracts, but we can see that the loans trend much lower for minority census tracts, closer to 100,000 dollars.

### Loading and cleaning banks dataset
```{r echo=FALSE, warning=FALSE, message=FALSE}
#loading banks data 
banks <- read.csv("banks.csv")

# replace the white space with underscores and lower-case all of the letters in column names
names(banks) <- names(banks) |> 
  str_replace_all("\\s", "_") |> 
  str_to_lower()
```
Replacing white space in the dataframe will help to make the dataset more readable with R programs.

```{r echo=FALSE, warning=FALSE, message=FALSE}

#filter the data to rows about banks that have valid longitudes and latitudes
banks_locations <- banks |>
  filter(!is.na(longitude), !is.na(latitude))

#convert longitude and latitude columns into spatial geometries
banks_locations <- st_as_sf(banks_locations, coords = c("longitude", "latitude"))

#set CRS to 4326 (WGS 84)
banks_locations <- st_set_crs(banks_locations, value = 4326)

#check the CRS
st_crs(banks_locations)

#plot map using ggplot2
banks_locations |>
  ggplot() +
  geom_sf(
    aes(color = name), alpha = 0.4, size = 2)

```
*** Say something about the locations/density of banks here ***


```{r}

# spatial join and census tracts 
#match CRS of both objects using the given CRS: 4326
banks_chloro <- st_transform(banks_locations, crs = st_crs(DMV_tracts))

#spatial join bank points and census tracts
banks_join <- st_join(DMV_tracts, banks_chloro, st_intersects) 

banks_merged_1 <- banks_join %>%
  group_by(GEOID10) %>%
  summarize(
    bank_count = sum(!is.na(name), na.rm = TRUE))

#join census tract back to new dataframe
banks_merged_1 <- st_join(DMV_tracts, banks_merged_1, join = st_intersects)

#trying to plot bank locations over a map of the census tracts 

#library(tmap)
#tm_shape(DMV_tracts) +
#tm_polygons()

#DMV_tracts_filter<- DMV_tracts %>%
 # select(GEOID10, geometry)

#plot(DMV_tracts_filter)

#tm_shape(DMV_tracts_filter) +
 # tm_polygons("banks_join")


# this map of bank location in DC works, but I'm not sure it's correct and it's not a choropleth
banks_join %>%
  ggplot()+ 
  geom_sf(data = DC_tracts) + 
  geom_sf(data = banks_locations, aes(color = ward)) 

#chloropleth

banks_merged_2 <- banks_merged_1 |>
  select(geometry, bank_count)

p1 <- banks_merged_2 %>%
  ggplot() +
  geom_sf(aes(fill = log1p(bank_count)), color = "white", size = 0.1) +
  scale_fill_gradient(low = "#81FFEF", high = "#F067B4", na.value = "white") +
  labs(fill = "Log(Bank Count)") +
  theme_void()

p1
  
```
# Banks Choropleth
```{r echo=FALSE, warning=FALSE, message=FALSE}

# trying to create a choropleth and it is not working lol
banks_merged <- banks_join %>%
  group_by(NAME10) %>%
  mutate(n = n()) %>%
  ungroup()
  
  
banks_merged_agg <- st_join(DC_tracts, banks_merged, join = st_intersects)  


# attempt 1
p1<- banks_merged_agg %>%
  
  ggplot() + geom_sf(data = DC_tracts) +
  geom_sf(aes(fill = n), size = 0.1) + coord_sf() +
  scale_x_continuous(breaks = 1:5) +
 scale_fill_gradient(
    low = "#90ee90", 
    high = "#F067B4",
    ) 
p1

library(scales)

p1<- banks_merged_agg %>%
  ggplot() + geom_sf(data = DC_tracts) +
  geom_sf(aes(fill = n), size = 0.1) + coord_sf() +
scale_fill_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(10, 12, 14, 16, 18, 20)),
                         guide = "colorbar")
p1

# attempt 2
banks_merged_try <- banks_join %>%
  group_by(NAME10) %>%
  mutate(n = n()) %>%
  ungroup()

banks_merged_try %>%
  ggplot() +
  geom_sf(aes(fill = n, color = n, alpha = 0.2)) 

```

# Geospatial Analysis
```{r echo=FALSE, warning=FALSE, message=FALSE}

# looking at economic outcomes in relationship to distance to banks 
banking<- st_transform(banks_join, crs = 4326)
DC<- st_transform(DC_tracts, crs = 4326)

econ <- debt_tract_2 %>%
  filter(COUNTYFP10 == "001") %>%
  st_as_sf(coords = c("INTPTLON10", "INTPTLAT10"), remove = FALSE) %>%
  st_set_crs(value = 4326) # setting CRS

banks_filtered<- st_join(banking, DC, left= FALSE)
banks_econ<- st_join(econ, st_buffer(banks_filtered, dist = 400))

max_banks_econ<- banks_econ %>%
  group_by(NAME10) %>%
  filter(DTI_ratio > 20) %>%
  summarize(n = n()) 

Banks_Debt <- max_banks_econ %>%
  ggplot() + 
  geom_sf(aes(fill = n)) + 
   scale_fill_gradient(
    low = "#D1E3F5",
    high = "blue",
    na.value = "white"
  )
```
*** Add description about bank/DTI ratio map here ***


```{r echo=FALSE, warning=FALSE, message=FALSE}

#looking at public health outcomes in relationship to distance to banks 

banking<- st_transform(banks_join, crs = 4326)
DC<- st_transform(DC_tracts, crs = 4326)

st_transform(Health_map, crs = 4326)
Health_map <- st_transform(Health_map, crs = st_crs(DMV_tracts))
Health_map <- st_join(DMV_tracts, Health_map, st_intersects) 


Health_DC <- Health_map %>%
  filter(COUNTYFP10 == "001") %>%
  filter(MeasureId == "DEPRESSION")

#healthyy <- st_as_sf(Health2, coords = c("longitude", "latitude"))
#healthyy <- st_set_crs(healthyy, value = 4326)
#healthyy<- st_transform(healthyy, crs = 4326)

#Filter depression
#healthyy <- healthyy %>%
 # filter(short_question_text %in% c("Depression"))
      

#transforming so it matches for the spatial join
banks_filtered<- st_join(banking, DC, left= FALSE)

st_crs(Health_DC)
banks_filtered <- st_transform(banks_filtered, crs = 4269)

banks_health<- st_join(Health_DC, st_buffer(banks_filtered, dist = 400))


max_banks <- banks_health %>%
  group_by(NAME10) %>%
  summarize(n = n())

view(max_banks)

Banks_Depression <- max_banks %>%
  ggplot() + 
  geom_sf(aes(fill = n)) + 
   scale_fill_gradient(
    low = "#D1E3F5",
    high = "blue",
    na.value = "white"
  )

Banks_Depression + Banks_Debt
```


## Creating visualizations from the debt-to-income dataset
Performing a merge with the tigris shapefiles to map debt-to-income ratios as a choropleth and interactive map
```{r echo=FALSE, warning=FALSE, message=FALSE}

debt_to_income$NAME10 <- as.numeric(debt_to_income$NAME10)

DMV_tracts$NAME10 <- as.numeric(DMV_tracts$NAME10)

debt_to_income$NAME10 <- debt_to_income$NAME10 / 100

#Merging the datasets by census tract name
debt_tract_2 <- full_join(DMV_tracts, debt_to_income, by = "NAME10")

debt_tract_2 %>%
  ggplot() + 
  geom_sf()

# Setting CRS for geospatial visualizations 

debt_tract_2 <- debt_tract_2 %>%
  st_as_sf(coords = c("INTPTLON10", "INTPTLAT10"), remove = FALSE) %>%
  st_set_crs(value = 4326) # setting CRS

st_crs(debt_tract_2) # Checking CRS

debt_tract_2 <- st_transform(debt_tract_2, crs = 4326)

st_crs(debt_tract_2)

# Debt-to-income ratio Choropleth
p3 <- debt_tract_2 %>%
  ggplot() + 
  geom_sf(
    aes(
      fill = DTI_ratio), color = "White", size = 0.1) +
  scale_fill_gradient(
    low = "#81FFEF",
    high = "#F067B4",
    na.value = "white"
  ) +
  labs(
    fill = "Debt to Income Ratio"
  ) +
  theme_void()

p3

```

This geospatial visualization demonstrates the range of debt-to-income ratios by census tract in the DMV. The range of colors provides a visualization of the range, but it would be helpful to know the averages for each census tract. 

### Using Leaflet to create an interactive map of debt to income ratios in the DMV
```{r echo=FALSE, warning=FALSE, message=FALSE}

# Load required libraries
library(leaflet)


# Create a new variable in our spatial object for fill color
debt_tract_2$FillColor <- colorQuantile("YlOrRd", debt_tract_2$DTI_ratio)(debt_tract_2$DTI_ratio)

# Create a leaflet map
leaflet(debt_tract_2) %>%
  
  # Add the polygons from the spatial object
  addPolygons(
    fillColor = ~FillColor,
    fillOpacity = 0.7,
    color = "white",
    weight = 1,
    popup = ~paste("Debt-to-Income Ratio: ", DTI_ratio)
  ) %>%
  
  # Add legend
  addLegend(
    "bottomright",
    pal = colorQuantile("YlOrRd", debt_tract_2$DTI_ratio),
    values = ~debt_tract_2$DTI_ratio,
    title = "Debt-to-Income Ratio",
    opacity = 1
  )

```
This map demonstrates the differences between the debt-to-income ratio for single family home purchases in the DMV area. You can click on each census tract to get specific information from each census tract.

## Visualizations of health outcomes in the DMV area
```{r echo=FALSE, warning=FALSE, message=FALSE}
#Merged CDC Places dataset with census tracts and made graph and choropleths

#Read in CSV of CDC data
Health1 <- read_csv("PLACES.csv")

#Clean data and get long and lat columns
Health2 <- Health1 %>%
  separate(col = Geolocation, into = c("point", "coordinates"), sep = " ", extra = "merge", remove = FALSE) |>
  separate(col = coordinates, into = c("longitude", "latitude"), sep = " ", remove = TRUE)

Health2 <- Health2 %>%
  mutate(longitude = gsub("[()]", "", longitude),
         latitude = gsub("[()]", "", latitude))

names(Health2) <- names(Health2) |> 
  str_to_lower()

#Use consistent health measures: CASTHMA, CHD, Depression, Diabetes, Stroke

Health3 <- Health2 %>%
  filter(measureid %in% c("CASTHMA", "CHD", "DIABETES", "STROKE", "DEPRESSION") &
           complete.cases(longitude, latitude))

#convert longitude and latitude columns into spatial geometries
sf_Health <- st_as_sf(Health3, coords = c("longitude", "latitude"))
#set CRS to 4326 (WGS 84)
sf_Health <- st_set_crs(sf_Health, value = 4326)
#check the CRS
st_crs(sf_Health)

#plot map using ggplot2
sf_Health |>
  ggplot() +
  geom_sf(
    aes(color = short_question_text), alpha = 0.4, size = 2) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Perform a Spatial Join and Create Choropleth

#match cRS
sf_Health2 <- st_transform(sf_Health, crs = st_crs(DMV_tracts))

#spatial_join health points and census tracts
Health_join <- st_join(sf_Health2, DMV_tracts, st_intersects)

#calculate the count of cancer and depression in new dataframe
health_merged <- Health_join |>
  group_by(NAME10) |>
  mutate(
    calculated_value = totalpopulation * (data_value / 100)
  )

#join census tract back to new dataframe
health_merged2 <- st_join(DMV_tracts, health_merged, join = st_intersects)

#create choropleth maps
<<<<<<< HEAD
p4 <- health_merged2 |>
  ggplot() +
=======
p4 <- health_merged |>
  ggplot() + geom_sf(data = DMV_tracts) +
>>>>>>> 5dfb9f7dbf8dd9bfa1a59c3b4be198170edcf5ef
  geom_sf(
    aes(
      fill = calculated_value), color = "blue", size = 0.1) +
  scale_fill_gradient(
    low = "#81FFEF", 
    high = "#F067B4",
    na.value = "white"
  ) +
  labs(
    fill = "Health Measures Count") +
  theme_void()

p4

combined_plot <- p3 + p4

combined_plot

```

#### Merging Health and Debt-to-Income dataset
```{r echo=FALSE, warning=FALSE, message=FALSE}

#Merging health dataset with debt_tract_2 

#Converting long and lat coordinates to sf objects
debt_tract_2_sf <- st_as_sf(debt_tract_2, coords = c("INTPTLON10", "INTPTLAT10"), crs = 4326)

# Convert 'Geolocation' column to an sf object
Health_sf <- st_as_sf(Health, wkt = "Geolocation", crs = 4326)

#Confirm if health and debt use the same coordinate reference systems (crs) 

#They do not. 
st_crs(debt_tract_2_sf)
st_crs(Health_sf)

#Transform debt dataset to match health
debt_tract_2_sf <- st_transform(debt_tract_2_sf, st_crs(Health_sf))

#Reconfirm if they have the same crs.
#Now they do.
st_crs(debt_tract_2_sf)
st_crs(Health_sf)

#Spatial Join
health_debt <- st_join(debt_tract_2_sf, Health_sf, join = st_intersects)

```

#### Machine Learning Model 

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidymodels)
set.seed(102102938)

split <- initial_split(data = debt_to_income, prop = 0.75)

debt_train <- training(x = split)
debt_test <- testing(x = split)

## Health outcome for machine learning model will be percentage of instances of depression in a censust tract

```

### Machine Learning Model #1: Predicting percentage of minorities in a census tract using the single family loan data
```{r echo=FALSE, warning=FALSE, message=FALSE}

debt_rec <- recipe(percent_minority ~ borrower_income + borrower_age + borrower_income_ratio + tract_income_ratio + DTI_ratio + interest_rate + note_amount, data = debt_train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_corr(all_numeric_predictors())

# Creating tuning grid for lasso model
lasso_grid <- grid_regular(penalty())

# LASSO model
lasso_mod <- linear_reg(penalty = tune(), mixture = 1) %>% 
  set_engine("glmnet")

# LASSO workflow
lasso_wf <- workflow() %>%
  add_recipe(debt_rec) %>%
  add_model(lasso_mod) 

folds <- vfold_cv(data = debt_train, v = 10, repeats = 1)

# LASSO fit
lasso_fit <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid = lasso_grid, 
    metrics = metric_set(rmse, mae), 
    control = control_grid(save_pred = TRUE))

# Evaluating models
lasso_metrics <- lasso_fit %>%
  collect_metrics(summarize = TRUE) 

lasso_fit %>%
  select_best(metric = "rmse")

# The RMSE of this model's best fit is around 22. 

```
FOR KAMIRYN: I'm honestly not sure I'm gonna leave this model in here, so feel free to skip over it when you're doing the write up. If I decide to keep it I will add something about it later. Thanks! However please include the second model (below) 

### Machine Learning Model #2: Predicting rates of depression in a census tract using economic predictors from the debt-to-income dataset

We've chosen a random forest model to model the relationship between economic factors and depression because random forest models adapt easily to many predictors, and nonlinear relationships. Below, we test two different random forest models and refine one with hyperparameter tuning to get a model with refined predictions of rates of depression in a census tract. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Filtering dataset for second machine learning  model

# We want only observations for rates of depression, and to exclude columns that have all values as NA or that are not interpretable by random forest modeling
depression_debt <- health_debt %>%
  drop_na(Data_Value) %>%
  filter(MeasureId == "DEPRESSION") %>%
  select(-Data_Value_Footnote, -Data_Value_Footnote_Symbol) %>%
  st_drop_geometry()

# Omit all NA values from dataset
depression_debt <- na.omit(depression_debt)

set.seed(12122023)
split2 <- initial_split(data = depression_debt, prop = 0.8)

depression_train2 <- training(x = split2)
depression_test2 <- testing(x = split2)


# Train the initial randomForest model with specified predictors
rf_mod <- randomForest(Data_Value ~ borrower_income + borrower_age + borrower_income_ratio + tract_income_ratio + DTI_ratio + interest_rate + note_amount + property_value + percent_minority, data = depression_train2)

rf_mod 

```
This model only explains about 77.6% of the variance in the rates of depression in each census tract. We think this is because there is a lack of tuning the model, and because we limited the number of predictors such that there was not enough information. 

After filtering out observations that are not interpretable by RF modeling, we will try a random forest model with all predictors

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Trying some tuning to see if we can find the best amount of mtrys to use for the model
bestmtry <- tuneRF(depression_train2,depression_train2$Data_Value,stepFactor = 2, improve = 0.01, trace = TRUE, plot = T, doBest = TRUE) 

# tuneRF shows that the best amount of mtrys is 32. Running random forest model again with 32 mtrys to see if it further optimizes the model. 
best_mtry <- 32

rf_mod3 <- randomForest(Data_Value ~., data = depression_train2, mtry = best_mtry)

rf_mod3

```

We used hyperparameter tuning to find the best number of mtrys for the model, and plotted the result. After finding that the best number of mtrys was 32, we added this parameter to the model. As a result, this model shows great improvement from the previous, explaining 99.99% of the variance. We can use this model to predict health outcomes on the testing dataset. 

```{r}

predicted_values <- predict(rf_mod3, newdata = depression_test2, type = "class")

test_predictions <- bind_cols(
  depression_test2, predicted_values)

test_predictions <- test_predictions %>%
  rename(.pred = ...99)

metrics(test_predictions, truth = Data_Value, estimate = .pred)

prediction_table <- tibble(test_predictions$Data_Value, test_predictions$.pred)
```
The prediction table shows that we have successfully trained a model that predicts 99% of the variance in our variable of interest, which reflects the rates of depression in each census tract. 

This model used predictors such as percent minority census tracts, debt to income ratio, median income of a census tract, and many other racial and economic factors in the data to be able to predict the depression rates in a census tract. 

Using a machine learning model that adapts to non-linear relationships has shown us that there is indeed a relationship between economic/racial factors and public health outcomes, such as depression. 

# Cluster Analysis
```{r echo=FALSE, warning=FALSE, message=FALSE}
# ATTEMPT

dti_data <- debt_to_income %>%
  select(NAME10, DTI_ratio)

#change NAME10 to categorical here

#maybe instead of NAME10 use percent_minority 

#try something that is less continuous, and has more discrete categories like county 

dti_pca_rec <- recipe(~ ., data = dti_data) %>%
  step_pca(all_numeric(), id = "pca") %>%
  prep(data = dti_data) %>%
  bake(new_data = dti_data) 


viz_1 <- dti_pca_rec %>%
  ggplot() +
  geom_point(
    mapping = aes(x = PC1, y = PC2),
    alpha = 0.5
  ) +
  theme_minimal()

viz_1 


#create the function 
debt_pca_fun<- function(DTI, df){
  debt_pca<- df %>%
    replace(is.na(.), 0) %>%
    filter(DTI_ratio == DTI) 
  
debt_pca_rec <- recipe(~ ., data = dti_data) %>%
    step_pca(all_numeric(), num_comp = 2) %>%
    prep(data = dti_data) %>%
    bake(new_data = dti_data) %>%
    select(PC1, PC2)
    

kmeans_rec<- recipe(~ ., data = dti_data) %>%
  step_select(all_numeric())
  
  debt_pca_spec <- k_means(
  num_clusters = 4
) %>%
  set_engine("stats", nstart = 100)

debt_pca_workflow <- workflow(
  preprocessor = kmeans_rec, 
  spec = debt_pca_spec
)

final_fit <- debt_pca_workflow %>%
  fit(data = dti_data)

debt_clusters <- 
  bind_cols(dti_data, debt_pca_rec, cluster = final_fit %>%
    extract_cluster_assignment() %>%
    pull(.cluster) )

return(debt_clusters)
}

#lastly, mapping the output
DTI<- 1:100
set.seed(20220413)
final_result <- map_dfr(.x = DTI, .f = debt_pca_fun, df = debt_to_income)
print(final_result)
  


```

```{r echo=FALSE, warning=FALSE, message=FALSE}
clust_plot <- final_result %>%
  ggplot() +
  geom_point(
    mapping = aes(x = PC1, y = PC2),
    alpha = 0.5
  ) +
  labs(
    title = "DTI",
    x = "PC1",
    y = "PC2"
  ) + 
  theme_classic()

clust_plot
```


```{r echo=FALSE, warning=FALSE, message=FALSE}

set.seed(20220414)
fviz_nbclust(
  dti_data, 
  FUN = hcut, 
  k.max = 10, 
  method = "wss", 
  hc_func = "hclust", 
  hc_method = "ward.D2", 
  hc_metric = "euclidean"
)

set.seed(20220414)
fviz_nbclust(
  dti_data,
  FUN = hcut,
  k.max = 10,
  method = "silhouette",
  hc_func = "hclust",
  hc_method = "ward.D2",
  hc_metric = "euclidean"
)

d <- dist(dti_data, method = "euclidean")
hclust_euc<- hclust(d, "ward.D2")
plot(hclust_euc, cex = 0.6, hang = -1, main = "Cluster Dendogram")
rect.hclust(hclust_euc, k = 4, border = "purple")

```
